---
title: <center> <h2> Spring 2024 - GE 461 Introduction to Data Science</h2> Dodgers Exploratory Data Analysis   </center>
pagetitle: 
author: <center> İsmail Kaan Özer 22002672, Görkem Kadir Solun 22003214</center>
always_allow_html: true
linkcolor: red
output: 
  bookdown::html_document2:
    theme: readable
    number_sections: false
    code_folding: "hide"
    toc: true
  bookdown::pdf_document2:
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RSQLite)
library(DBI)
library(tidyverse)

con <- dbConnect(SQLite(),"data/dodgers.sqlite")
dbListTables(con)

tbl(con, "events") %>% 
  collect() %>% 
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
         month = factor(month, levels = c("APR","MAY","JUN","JUL","AUG","SEP","OCT"))) %>% 
  mutate_if(is.character, factor) %>% 
  mutate(temp = round((temp- 32)*5/9)) -> events
  
```

# Introduction

Los Angeles Dodgers Baseball team gathered data of their 81 games to analyze. Objective is increasing the attendance to the matches.  Temperature and Attendance variables are quantitative while all others are qualitative. Specific analysis target is measuring the effects of all variables on attendance variable. Below is the summary of the data set.

```{r}
summary(events)
```

# Visualizing and Evaluating Different Variables

In order to observe whether different variables increase attendance, Chi-squared tests and Welch’s Two Sample t-tests was conducted on variables and attendance. 

## Booblehead

Firstly, we need to visualize the attendance with booblehead promotion existence, then we can decide whether there was a correlation between the attendance and boblehead promotions.

```{r}
events %>% 
  ggplot(aes(bobblehead, attend)) +
  geom_boxplot()
```

Graph suggests that there is a difference. However, to be sure, Welch Two Sample t-test conducted between attendance and booblehead promotions.

```{r}
t.test(x=events$attend[events$bobblehead=="YES"], y=events$attend[events$bobblehead=="NO"])
```

A statistical test (t-test) was done, and the results strongly suggest that bobblehead promotions have a positive impact on attendance. The test showed a clear connection between promotions and higher attendance numbers. We are 95% confident that bobblehead promotions increase attendance by somewhere between 11,500 and 16,500 people per event.

A Chi-squared test also conducted to check whether there is a connection between between booblehead promotions and higher attendance numbers.

```{r}
events %>% 
  group_by(bobblehead) %>% 
  summarize(avgattend = mean(attend)) %>% 
  pull(avgattend) %>% chisq.test()
```
Similar to previous test, the results strongly suggest that bobblehead promotions have a positive impact on attendance. The test showed a clear connection between promotions and higher attendance numbers. Our p-value is near zero meaning we are confident that bobblehead promotions increase attendance.

## Fireworks

Same as booblehead, firstly, we need to visualize the attendance with fireworks existence, then we can decide whether there was a correlation between the attendance and fireworks.

```{r}
events %>% 
  ggplot(aes(fireworks, attend)) +
  geom_boxplot()
```

In contrast to booblehead, graph suggests that there is no difference. However, to be sure, Welch Two Sample t-test conducted between attendance and fireworks.

```{r}
t.test(x=events$attend[events$fireworks=="YES"], y=events$attend[events$fireworks=="NO"])
```

A statistical test (t-test) was done, and the results suggest that fireworks have little to no impact on attendance. It is seen that there is enough evidence to conclude that there isn’t a relation between attendance and fireworks considering the p-value of test is almost 1. The test showed a no connection between fireworks and higher attendance numbers.

To be sure, a Chi-squared test also conducted to check whether there is a connection between between fireworks and higher attendance numbers.

```{r}
events %>% 
  group_by(fireworks) %>% 
  summarize(avgattend = mean(attend)) %>% 
  pull(avgattend) %>% chisq.test()
```

Similar to previous test, the results suggest that bobblehead promotions have little to no impact on attendance. The test showed not a clear connection between fireworks and higher attendance numbers. Our p-value is near one meaning we are confident that fireworks not related to attendance.

## Shirt

Firstly, we need to visualize the attendance with shirt promotion existence, then we can decide whether there was a correlation between the attendance and shirt promotions.

```{r}
events %>% 
  ggplot(aes(shirt, attend)) +
  geom_boxplot()
```

Graph suggests that there is a difference. However, again, to be sure, Welch Two Sample t-test conducted between attendance and shirt promotions.

```{r}
t.test(x=events$attend[events$shirt=="YES"], y=events$attend[events$shirt=="NO"])
```

A statistical test (t-test) was done, and the results suggest that shirt promotions have no clear impact on attendance as p-value is higher than 0.05 so we accept the null and conclude that there is a not much difference. The test showed a no concise connection between shirt promotions and higher attendance numbers.

A Chi-squared test also conducted to check whether there is a connection between between shirt promotions and higher attendance numbers.

```{r}
events %>% 
  group_by(shirt) %>% 
  summarize(avgattend = mean(attend)) %>% 
  pull(avgattend) %>% chisq.test()
```
In contrast to previous test, the results strongly suggest that shirt promotions have a positive impact on attendance. The test showed a connection between shirt promotions and higher attendance numbers. Our p-value is near zero meaning we are confident that shirt promotions increase attendance.

Even though the two p-values appear contradictory, they reveal different aspects of data:

- The t-test suggests that those who got a shirt and those who didn't might have similar average attendance levels.
- However, the Chi-squared test reveals that there's an important relationship between attendance and receiving a shirt. This might mean that people with higher attendance have a higher chance of getting a shirt, even if the mean differences are subtle.

Therefore, we will be considering shirts relation in attendance.





























```{r}
lm1 <- lm(attend ~ ., events)
summary(lm1)
```


```{r}
anova(lm1)
```

```{r}
dbDisconnect(con)
```

