---
title: <center> <h2> Spring 2024 - GE 461 Introduction to Data Science</h2> Dodgers Exploratory Data Analysis   </center>
pagetitle: 
author: <center> İsmail Kaan Özer 22002672, Görkem Kadir Solun 22003214</center>
always_allow_html: true
linkcolor: red
output: 
  bookdown::html_document2:
    theme: readable
    number_sections: false
    code_folding: "hide"
    toc: true
  bookdown::pdf_document2:
    number_sections: false
knitr: opts_chunk$set(warning = FALSE)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RSQLite)
library(DBI)
library(tidyverse)
library(car)
library(pander)

con <- dbConnect(SQLite(),"data/dodgers.sqlite")
dbListTables(con)

tbl(con, "events") %>% 
  collect() %>% 
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
         month = factor(month, levels = c("APR","MAY","JUN","JUL","AUG","SEP","OCT"))) %>% 
  mutate_if(is.character, factor) %>% 
  mutate(temp = round((temp- 32)*5/9)) -> events
  
```

# Introduction

The Los Angeles Dodgers Baseball team gathered data from their 81 games to analyze. The objective is to increase attendance at the matches. Temperature and Attendance variables are quantitative, while all others are qualitative. The specific analysis target is measuring the effects of all variables on the attendance variable. Below is the summary of the data set.

```{r}
summary(events)
```

# Visualizing and Evaluating Different Variables

To observe whether different variables increase attendance, Chi-squared tests, f-tests, and Welch’s Two Sample t-tests were conducted on variables and attendance. To make sure of the result of analysis, we define a linear regression model with all variables to compare with different models afterwards.

```{r}
lm1 <- lm(attend ~ ., events)
```

## Bobblehead

Firstly, we need to visualize the attendance with bobblehead promotion existence; then, we can decide whether there was a correlation between the attendance and bobblehead promotions.

```{r}
events %>% 
  ggplot(aes(bobblehead, attend)) +
  geom_boxplot()
```

The graph suggests that there is a difference. However, to be sure, a Chi-squared test was conducted between attendance and bobblehead promotions.

```{r}
events %>% 
  group_by(bobblehead) %>% 
  summarize(avgattend = mean(attend)) %>% 
  pull(avgattend) %>% chisq.test()
```
Our p-value is near zero, so we can say that bobblehead promotions will increase attendance. Similar to the previous test, the results strongly suggest that bobblehead promotions positively impact attendance. The test showed a clear connection between promotions and attendance numbers.

We can also see how many fans are expected to attend by a bobblehead promotion.

```{r}
confint(lm1, parm = "bobbleheadYES", level = 0.95)
```

Almost 15850 fans are expected in the game if bobblehead promotion exists.

Lastly, we define a linear regression model without bobblehead variable and apply F-test to fully understand that bobblehead is important.

```{r}
model_wout_bobblehead <- lm(attend ~ month + day + day_of_week + opponent + temp + skies + day_night + cap + shirt + fireworks, data = events)
anova(lm1, model_wout_bobblehead)
```
p-value resulted less than 0.01: A low p-value suggests that there is a significant difference between the two models because null hypothesis suggests that there is no difference. Since there is a significant difference between two models, we can not ignore bobblehead variable because it is important for our model. Now, we are sure that bobblehead promotions increase the attendance.

## Fireworks

Like the bobblehead, firstly, we need to visualize the attendance with the fireworks' existence, and then we can decide whether there was a correlation between the attendance and fireworks.

```{r}
events %>% 
  ggplot(aes(fireworks, attend)) +
  geom_boxplot()
```

In contrast to bobbleheads, the graph suggests that there is no difference. However, to be sure, the Welch Two Sample t-test and Chi-squared test was conducted between attendance and fireworks.

```{r}
t.test(x=events$attend[events$fireworks=="YES"], y=events$attend[events$fireworks=="NO"])
```

A statistical test (t-test) was done, and the results suggest that fireworks have little to no impact on attendance. It is seen that there is enough evidence to conclude that there isn’t a relation between attendance and fireworks, considering the p-value of the test is almost 1. The test showed no connection between fireworks and attendance numbers.

A Chi-squared test was also conducted to double check whether there is a connection between fireworks and attendance numbers.

```{r}
events %>% 
  group_by(fireworks) %>% 
  summarize(avgattend = mean(attend)) %>% 
  pull(avgattend) %>% chisq.test()
```

Similar to the previous test, the results suggest that firework shows have little to no impact on attendance. The test showed no apparent connection between fireworks and attendance numbers. Our p-value is near one, meaning we are confident that fireworks are unrelated to attendance.

## Shirt

Firstly, we need to visualize the attendance with shirt promotion existence, and then we can decide whether there is a correlation between attendance and shirt promotions.

```{r}
events %>% 
  ggplot(aes(shirt, attend)) +
  geom_boxplot()
```

The graph very strongly suggests that there is a difference. However, again, to be sure, a Chi-squared test was also conducted to check whether there is a connection between shirt promotions and attendance numbers.

```{r}
events %>% 
  group_by(shirt) %>% 
  summarize(avgattend = mean(attend)) %>% 
  pull(avgattend) %>% chisq.test()
```
The results strongly suggest that shirt promotions positively impact attendance. The test showed a connection between shirt promotions and attendance numbers. Our p-value is near zero, so we are confident that there is difference and shirt promotions will increase attendance.

The Chi-squared test reveals an important relationship between attendance and receiving a shirt. This might mean that people with higher attendance have a higher chance of getting a shirt, even if the mean differences are subtle.

Therefore, we will focus on our results in Chi-squared tests and consider shirt relations in attendance.

## Cap

Firstly, we need to visualize the attendance with the existence of cap promotions, and then we can decide whether there is a correlation between attendance and cap promotions.

```{r}
events %>% 
  ggplot(aes(cap, attend)) +
  geom_boxplot()
```

The graph suggests that there is a very small difference. However, again, to be sure, a Welch Two Sample t-test was also conducted to check whether there is a connection between cap promotions and attendance numbers.

```{r}
t.test(x=events$attend[events$cap=='YES'],
       y=events$attend[events$cap=='NO'])
```
The results suggest that cap promotions have a near to zero impact on attendance. The test showed a disconnection between cap promotions and attendance numbers. Our p-value is more than 0.05, so we are confident that cap promotions is not related to attendance. To make completely sure, anova test is applied.

```{r}
model_wout_cap <- lm(attend ~ month + day + day_of_week + opponent + temp + skies + day_night + shirt +  fireworks + bobblehead, data = events)
anova(lm1, model_wout_cap)
```
Since p-value is higher than 0.05, we are sure that cap promotions are not related to attendance.
Therefore, we will not consider cap relations in attendance.

## Day and Night

Firstly, we need to visualize the attendance day and night; then, we can decide whether there was a correlation between the attendance and day or night.

```{r}
events %>% 
  ggplot(aes(day_night, attend)) +
  geom_boxplot()
```

The graph suggests that there is very small to no difference. However, again, to be sure, a Welch Two Sample t-test was conducted to check whether there is a connection between day or night and attendance numbers.

```{r}
t.test(x=events$attend[events$day_night=='Day'],
       y=events$attend[events$day_night=='Night'])
```
Like the graph, the results suggest that being day or night has no apparent impact on attendance. The test showed a disconnection between being day or night and attendance numbers. Our p-value is more than 0.05, meaning we are confident that being day or night is not related to attendance. Lastly, anova test is applied.

```{r}
model_wout_time <- lm(attend ~ month + day + day_of_week + opponent + temp + skies + shirt + cap +  fireworks + bobblehead, data = events)
anova(lm1, model_wout_time)
```
Anova test result suggests that time that game is played does not have a relation with attendance because p value is greater than 0.05.
Therefore, we will not consider day and night relations in attendance.

## Skies

Firstly, we need to visualize the attendance with skies; then, we can decide whether there was a correlation between the attendance and sky.

```{r}
events %>% 
  ggplot(aes(skies, attend)) +
  geom_boxplot()
```

The graph suggests that there is very small difference. However, again, to be sure, a Chi-squared test was conducted to check whether there is a connection between sky and attendance numbers.

```{r}
events %>% 
  group_by(skies) %>% 
  summarize(avgattend = mean(attend)) %>% 
  pull(avgattend) %>% chisq.test()
```
Somewhat similar to the graph, the results suggest that skies have a impact on attendance. The test showed a connection between sky and attendance numbers. Our p-value is near zero meaning we are confident that sky is related to attendance.

Again, to be sure, a Welch Two Sample t-test was conducted to check whether there is a connection between sky and attendance numbers.

```{r}
t.test(x=events$attend[events$skies=='Clear'],
       y=events$attend[events$skies=='Cloudy'])
```

In contrast to the Chi-squared test, the results suggest that skies have a low impact on attendance. The test showed a disconnection between sky and attendance numbers. Our p-value is more than 0.05, which means we are not totally confident that sky is related or unrelated to attendance. We can be sure after applying F-test.

```{r}
model_wout_skies <- lm(attend ~ month + day + day_of_week + opponent + temp + day_night + cap + shirt + fireworks + bobblehead, data = events)
anova(lm1, model_wout_skies)
```
F test's p value is very close to 1. This result means that two models are not significantly different. So, skies is not an important variable.
Therefore, we will not consider skies relations in attendance.

## Temperature

Firstly, we need to visualize the attendance with temperature; then, we can decide whether there was a correlation between the attendance and temperature.

```{r}
events %>% 
  ggplot(aes(temp, attend)) +
  geom_point() +
  geom_smooth(se=FALSE)
```

The graphs suggests a difference, particularly around 22-23 degrees. We can see higher averages of attendance. Therefore, we will consider temperature relations in attendance.

### Temperature with other variables

```{r}
events %>% 
  ggplot(aes(temp, attend)) +
  geom_jitter(aes(col = opponent)) +
  geom_text(aes(label = str_sub(opponent, 1,4), col = opponent)) +
  geom_smooth(se = FALSE)
```

Here, we displayed the opponents, temperature, and attendance. Many variances exist, so we should explore this further by considering these variables.

```{r}
events %>% 
  ggplot(aes(temp, attend)) +
  geom_jitter(aes(col = bobblehead)) +
  geom_smooth(se = FALSE)
```

Here, we can see how the day and night temperature varies through the months.

```{r}
events %>% 
  ggplot(aes(temp, attend)) +
  geom_jitter(aes(col = day_of_week)) +
  geom_text(aes(label = str_sub(day_of_week, 1,3), col = day_of_week)) +
  geom_smooth(se = FALSE)
```

Temperature and days of the week of the games played against attendance.

```{r}
events %>% 
  ggplot(aes(temp, attend)) +
  geom_jitter(aes(col = month)) +
  geom_text(aes(label = str_sub(month, 1,3), col = month)) +
  geom_smooth(se = FALSE)
```

Temperature and months of the games played against attendance.

```{r}
events %>% 
  ggplot(aes(temp, attend)) +
  geom_jitter(aes(col = bobblehead)) +
  geom_smooth(se = FALSE)
```

As an another analysis for temp variable, we got a hint from the plot: our shape looks like a -x^2 line. We can play with it to make it more related. Below is the different lines for temp and BIC values of them.

```{r}
model1<-lm(attend~temp,events)
model2 <- lm(attend~I(temp^2),events)
model3 <- lm(attend~I(temp^2) + I(temp^3),events)
model4 <- lm(attend~I(temp^2) + I(temp^3)+I(temp^4),events)
model5<- lm(attend~poly(temp,2),events)
model6<- lm(attend~poly(temp,3),events)
model7<- lm(attend~poly(temp,4),events)
BIC(model1,model2,model3,model4,model5, model6,model7)
```

The model5 has the least BIC value, which is temp^2. Let's test that model.

```{r}
summary(model5)
```

As you can see, p-value is less than the indicated level. Now, temp is related to attendance rate. 
Therefore, we will consider poly(temp,2) in attendance.

## Days and Month

Firstly, we need to visualize the attendance with dates; then, we can decide whether there was a correlation between the attendance.

```{r}
events %>%
mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
count(day_of_week, name = "Number of games") %>%
rename(`Week day`= day_of_week) %>%
pander(caption = "Number of games on week days")
```

The games were played uniformly every weekday except Thursday, which has less than other days. 

```{r}
events %>%
mutate(month = factor(month, levels = c("APR", "MAY", "JUN", "JUL", "AUG", "SEP", "OCT"))) %>%
count(month, name = "Number of games") %>%
rename(`Month`= month) %>%
pander(caption = "Number of games across months")

```

May hosted the most significant number of games, while October the least. June has as much as half of the games in May. The remainder of the months have high and similar game numbers.

This heatmap shows total attendance versus weekday and month. The colors change from bright yellow to dark red as attendance increases. Default heatmap shuffles rows and columns to bring together weekdays and months with similar attendance. Here, we see May, Aug, and Jul together within the months and Saturday, Friday, and Sunday within the weekdays.

```{r}
xtabs(attend ~ day_of_week + month, events) %>% 
  heatmap()
```

The heatmap suggests a difference. For example, attendance is low in October, and we can see there is higher attendance on Monday, Tuesday, and Wednesday in July. Still, on other days, attendance is low. Therefore, we will consider dates, months, and days relations in attendance. We can use one-way anova test to double check.

```{r}
anova(lm1)
```
P-values of both month and day_of_week suggests that they are important.
We can also draw a graph to see how attendance varies across the days of the week.

```{r}
events %>% 
  ggplot(aes(day_of_week, attend)) +
  geom_boxplot()
```

Here, we can see that we see a higher average on Tuesdays and a lower average on Mondays. It can be seen in summary of our model with all variables.

```{r}
lm1 <- lm(attend ~ ., events)
summary(lm1)
```

Only day_of_week Tuesday's p-value is less than the wanted level. 

## Opponents

Firstly, we need to visualize the attendance with opponents; then, we can decide whether there was a correlation between the attendance and opponents.

```{r}
events %>% 
  ggplot(aes(opponent, attend)) +
  geom_boxplot()
```

The graph suggests that there is difference. However, again, to be sure, a Chi-squared test was conducted to check whether there is a connection between opponents and attendance numbers.

```{r}
events %>% 
  group_by(opponent) %>% 
  summarize(avgattend = mean(attend)) %>% 
  pull(avgattend) %>% chisq.test()
```
Same as the graph, the results suggest that opponents have a impact on attendance. The test showed a connection between opponents and attendance numbers. Our p-value is near zero meaning we are confident that opponent is related to attendance. We should check other tests.

```{r}
model_wout_opponent <- lm(attend ~ month + day + day_of_week + temp + skies + day_night + cap + shirt + fireworks + bobblehead, data = events)
anova(lm1, model_wout_opponent)
```
Contrary to graph and chi-sq test, anova test suggests that oppenents is not an important variable.
```{r}
summary(lm1)
```
As it is seen in summary, there is not any opponent row that suggests opponents is an important variable.
Therefore, we will not consider opponents relations in attendance.

# Final Model and Validating

We need to examine which criteria in the dataset affected attendance to ensure that variables do not coincide with other factors in a way that could skew this test. To achieve this, a multiple regression model was developed that included all criteria and removed them one at a time until the results stopped improving. The model was trained using a random selection of 85% of the dataset; the remaining games were utilized to assess the accuracy of the predictions.

```{r}
set.seed(666)
sampleSize <- floor(0.85 * nrow(events))

train_ind <- sample(seq_len(nrow(events)), size = sampleSize)

train <- events[train_ind, ]
test <- events[-train_ind, ]

model <- lm(formula = attend ~ month + day + day_of_week + opponent + temp + skies + day_night + cap + shirt + fireworks + bobblehead, data = train)

summary(model)
```

Here, we included all variables to start our improvement of the final model.

The F-statistics p-value is less than 0.05, and then there is a relation between attendance and predictors.

The model is tested and are compared using the Mean Absolute Error. A train sample and a test sample were randomly selected from the dataset using the same seed, ensuring consistency across all model tests.

```{r}
prediction <- predict(model,test)

mean(abs(prediction-test$attend))
```

The tests done between the variables and attendance will be considered the first model to check whether there is a correlation between the two in order to expedite the variable removal process. The accuracy of the model should theoretically increase if a variable is eliminated if it does not correspond with attendance. By removing our unrelated values, we get a final model.

The Mean Absolute Error calculation of the test.

```{r}
final <- lm(formula = attend ~ month + day + day_of_week 
             #+ opponent
             + poly(temp,2) 
             #+ skies 
             #+ day_night
             #+ cap
             + shirt 
             #+ fireworks
             + bobblehead
             , data = train)

finalpred <- predict(final,test)

mean(abs(finalpred-test$attend))
```

We see a drop in the MAE of the predictions when the unrelated variables are removed from the initial regression model. These variables are skies, opponent, day or night, cap promotions, and fireworks. Our models prediction is better considering the MAE calculation.

```{r}
summary(final)
```

The final model yields the best result. Looking at the p-values, we can see here that included variables (day of the week, month, bobblehead, shirt, temp^2) have an effect. 

We will compare our final models diagnostic plots with diagnostic plots of our previous model which has all variables.

Plots of our previous model

```{r, warning=FALSE}
model <- lm(formula = attend ~ month + day + day_of_week + opponent + temp + skies + day_night + cap + shirt + fireworks + bobblehead, data = events)
plot(model)
```

Our final models plot. We can see diagnostics of our final model here.

```{r}
final <- lm(formula = attend ~ month + day + day_of_week 
             #+ opponent
             + poly(temp,2) 
             #+ skies 
             #+ day_night
             #+ cap
             + shirt 
             #+ fireworks
             + bobblehead
             , data = events) # make it with full data

plot(final)
```

Interpretations of the comparison

- Fitted vs Resiudals plot has a better shape than the previous model. The red curve is around 0. But still somehow has a U kind of shape but it is very small compared to previous.
- In the final Q-Q Residuals plot there are more points closer to the line than the previous one. The final is more closer to 45 degree line.
- The final Scale-Location plot indicates a relatively equal spread of residuals across all fitted values, satisfying homoscedasticity.
- The final Cook's Distance values are less than 0.5. Hence there are no influential points based on this criteria.

Fitted value plots of our previous model

```{r}
events %>% 
  mutate(fitted = fitted(model)) %>% 
  ggplot(aes(fitted, attend)) +
  geom_point() +
  geom_abline() +
  geom_smooth(se=FALSE)
```

Fitted value plots of our final model

```{r}
events %>% 
  mutate(fitted = fitted(final)) %>% 
  ggplot(aes(fitted, attend)) +
  geom_point() +
  geom_abline() +
  geom_smooth(se=FALSE)
```

Here, we can see our final model has fewer diverted points to the standard line.

With the best multiple regression model to predict attendance, including bobblehead promotion, shirt promotion, hotter months, Tuesdays, and temperatures around 23-24, we can see that these variables contribute positively to the games' attendance, considering the tests conducted and models produced earlier.

```{r}
dbDisconnect(con)
```