---
title: <center> <h2> Spring 2024 - GE 461 Introduction to Data Science</h2> Dodgers Exploratory Data Analysis   </center>
pagetitle: 
author: <center> İsmail Kaan Özer 22002672, Görkem Kadir Solun </center>
always_allow_html: true
linkcolor: red
output: 
  bookdown::html_document2:
    theme: readable
    number_sections: false
    code_folding: "hide"
    toc: true
  bookdown::pdf_document2:
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RSQLite)
library(DBI)
library(tidyverse)

con <- dbConnect(SQLite(),"data/dodgers.sqlite")
dbListTables(con)

tbl(con, "events") %>% 
  collect() %>% 
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
         month = factor(month, levels = c("APR","MAY","JUN","JUL","AUG","SEP","OCT"))) %>% 
  mutate_if(is.character, factor) %>% 
  mutate(temp = round((temp- 32)*5/9)) -> events
  
```

# Introduction

Los Angeles Dodgers Baseball team gathered data of their 81 games to analyze. Objective is increasing the attendance to the matches.  Temperature and Attendance variables are quantitative while all others are qualitative. Specific analysis target is measuring the effects of all variables on attendance variable. Below is the summary of the data set.

```{r}
summary(events)
```


```{r}
lm1 <- lm(attend ~ ., events)
summary(lm1)
```


```{r}
anova(lm1)
```
